Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-03-02 Manuscript/manuscript.tex
Started analysis of clusrank code

\section{2019 and National Data}
This problem of whether or not the sensor was faulty can be further explored in
two ways. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world class athletes to not improve significantly from one year to the
next, but that is not neccessarily the results that were found.

@@

The model accuracy improves when certain batches are removed from the data.  Batch 41 in
particular has extremely high reaction times.  This outlier can likely be explained by a
few different causes.  Batch 41 is a semifinals heat of the 2013 World Championship.  2013 had the highest average reaction time from the pooled data set.  2013 was also the second world championship after the rule change that imposed the one strike policy on
disqualifications. 2011 and 2013 had the highest average reaction time in the pooled
data and were the two years after the rule change.  What this suggests is that there could have been a psychological effect of the stricter policy that resulted in athletes being more cautious and reacting slower than they had in the previous two years under the more lax guidelines from 2007 to 2009 in order to avoid a false start and thus a disqualification.
In Batch 41 particularly two runners did not finish, and two runners reacted extremely slow: 0.327 and 0.397 (Remember that Devon Allen's reaction time was
0.099 seconds in the 2022 Championship final and 0.101 in the 2022 semifinal).  Removing this batch and others could greatly improve the strength of the model.  However, while
it is a high outlier, there is not sufficient information available as to why the
reaction times are so high, and no basis for why it should be improved other than because
it is an outlier.

@@

The figure below plots the average reaction time by year for the pooled men's data since
1999.  The circle in the bottom right is for 2022, which is signifcantly lower than the
data points for the previous three years, and is substially lower than the data points
for 2011 and 2013.  This graph supports the conclusion that the timing system from 2022
had some sort of error that caused the reaction times to be so much lower.  Assuming there
was not an issue with the system, the large variation is confusing because there is little explanation
as to why there is so much of a drop off in reaction time from 2013 to 2015 or the spike
from 2001 to 2003.

Devon Allen's disqulification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Addtionally, it was shown
through a gamma mixed effects model that there are significant year and batch
effects that greatly impact reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not neccessarily
be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
second barrier again at the next World Track Championship (August 2023), then World Athletics 
needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
reflexes are not penalized like how Allen was in 2022.  World Athletics and Seiko 
may want to do further testing of their own to determine what is a human's physically
fastest reaction time, and if they choose to do so, they should start with Devon
Allen.

>>>>

\section{2019 and National Data}
This problem of whether the sensor was faulty can be further explored in
two ways. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next, but that is not necessarily the results that were found.  The R package
"clusrank" was used to perform a clustered Wilcoxon signed-rank test.  This was
chosen over a typical t test so as to avoid any assumptions about the distribution
of population means.  The clusrank package allows clusters to be formed, which
in this case are for each athlete.  The data containing each athlete's name,
nationality, and reaction times for the 2019 and 2022 World Championship can be
found in the csv file labeled "2019vs2022".  This data was transformed to "Clusrank2019vs2022"
so that it better fits the clusrank package.  Each athlete was given a non-meaningful
ID (An ID of 1 does not indicate that this athlete is better or worse than any
other athlete).  Additionally, the "Year" column is used to determine what year
the time was registered with "1" indicating 2019 and "2" indicating 2022.

Using the "clusWilcox.test" function, a paired and clustered Wilcoxon signed-rank
test was performed wtih the cluster set to the be the Athlete and group set to Year.
Two different methods were used: "rgl" which is (Prof Yan should help with this)
Paired was set to be "true" to perform a matched pairs test, and the alternative
hypothesis was set to be "greater" as we want to know if the 2019 reaction times
were larger than the 2022 times by a significant margin.

When the clusrank analysis was performed, the "ds" method produced a z score of
3.5271 and p value of 0.0002101 and the "rgl" method produced a z score of 3.665
and a p value of 0.0001237.  These values are for a one-sided test and show that
the mean reaction time on average in 2019 was larger than the mean reaction time
in 2022 for athletes who competed in both championships.

Another important data set that was investigated was the 2022 national competition
data.  In order to send between 1-4 athletes to the World Championships, almost
every athlete had to qualify by proving they are elite at the national level.
Thus, data was compiled on athletes who competed at the 2022 World Track and 
Field Championships on what these athletes also ran at their country's national
championship.  If there was any argument that athletes have improved their reaction
time from 2019 to 2022, it should now be considered nearly null, as the national
competitions took place between June and July 2022 and the World Championships
were in August.  The equipment used to measure reaction time cannot be guaranteed
to be the same from national to international competition, but the time span between
meets has been reduced considerably.  Repeating the data collection steps and
again performing a clustered analysis for a matched pairs Wilcoxon signed-rank test
results in two new z scores and p values.  The "ds" method produced a z score of
3.6755 and p value of 0.0001187 and the "rgl" method produced a z score of 3.7618
and a p value of 8.434e-05. These results are highly significant and show that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships.

@@

The model accuracy improves when certain batches are removed from the data.  Batch 41 in
particular has extremely high reaction times.  This outlier can likely be explained by a
few different causes.  Batch 41 is a semifinals heat of the 2013 World Championship.  2013 had the highest average reaction time from the pooled data set.  2013 was also the second world championship after the rule change that imposed the one strike policy on
disqualifications. 2011 and 2013 had the highest average reaction time in the pooled
data and were the two years after the rule change.  What this suggests is that there could have been a psychological effect of the stricter policy that resulted in athletes being more cautious and reacting slower than they had in the previous two years under the more lax guidelines from 2007 to 2009 in order to avoid a false start and thus a disqualification.
In Batch 41 particularly two runners did not finish, and two runners reacted extremely slow: 0.327 and 0.397 (Remember that Devon Allen's reaction time was
0.099 seconds in the 2022 Championship final and 0.101 in the 2022 semifinal).  Removing this batch and others could greatly improve the strength of the model.  However, while
it is a high outlier, there is insufficient information available as to why the
reaction times are so high, and no basis for why it should be improved other than because
it is an outlier.

@@

The figure below plots the average reaction time by year for the pooled men's data since
1999.  The circle in the bottom right is for 2022, which is significantly lower than the
data points for the previous three years, and is substially lower than the data points
for 2011 and 2013.  This graph supports the conclusion that the timing system from 2022
had some sort of error that caused the reaction times to be so much lower.  Assuming there
was not an issue with the system, the large variation is confusing because there is little explanation
as to why there is so much of a dropoff in reaction time from 2013 to 2015 or the spike
from 2001 to 2003.

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and batch
effects that greatly impact reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not necessarily
be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
second barrier again at the next World Track Championship (August 2023), then World Athletics 
needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
reflexes are not penalized like how Allen was in 2022.  World Athletics and Seiko 
may want to do further testing of their own to determine what is a human's physically
fastest reaction time, and if they choose to do so, they should start with Devon
Allen.
